---
title: "Comparison of CARD-FISH and DNA tag sequencing in Muskegon Lake"
author: "Edna Chiang"
date: "October 22, 2014"
output:
  html_document:
    keep_md: yes
---



#### Introduction
  For my senior honors thesis, I am studying the distribution and lifestyle of the bacterial phylum *Verrucomicrobia* in the Great Lakes. Species of *Verrucomicrobia* have been found to be involved in complex carbon breakdown as well as methane oxidation, which suggest that they play a role in the carbon cycle^3,4^. The majority of *Verrucomicrobia* studies have been conducted in soil where they are a ubiquitous and abundant group5. They were first observed and later isolated from aquatic systems^6,7,8^; however, very few freshwater studies have been conducted and much remains to be learned regarding their ecology, niche, and function in the ecosystem^9^.
  
  The majority of my data will be generated from DNA sequencing; however, due to the potential for bias in the molecular data, I will complement it with catalyzed reporter deposition in-situ hybridization (CARD-FISH), analysis to generate the most accurate results possible. I spent the past year optimizing CARD-FISH, and will now compare it to the molecular data. My question is whether the proportion of *Verrucomicrobia* reported by the two methods are similar.


#### Methods
  Samples were collected during a sampling cruise in May 2014 conducted in collaboration with the Grand Valley State University Annis Water Resources Center. 2 L of water was collected in Van Dorn Water Samplers at varying depths in Muskegon Lake in three locations: the river inlet, the estuary outlet, and the deepest point of the lake. The water was first pre-filtered through 210 μm and 20 μm to remove debris. 20 mL of this filtrate was collected and stored on ice or in a 4°C refrigerator for less than 24 hrs before being filtered back in the lab. CARD-FISH was performed on these filters, according to the method of Sekar, et al. Probes Eub 338, EUB 338 II, and EUB 338 III were used to tag bacterial cells. *Verrucomicrobia* was targeted by mixing EUB 338 II and unlabeled EUB 338 III. The nonsense probe NONEUB (NON338) was used as a control. Filters were counterstained with DAPI to identify all cells. The filters were then examined with fluorescence microscopy, and a minimum of 1000 cells was counted per labeled filter.
  
  
   The remaining pre-filtered was pumped through a 3 μm and 0.22 μm filter, which, respectively, captured particle-associated and free-living microorganisms. These filters were stored in liquid nitrogen and, upon returning to the lab, in an -80°C refrigerator. The Qiagen AllPrep DNA Extraction Kit was used to extract DNA from the filters. Subsequently. The 16S rRNA V4 region was tag-sequenced at the University of Michigan, and analyzed using Mothur and RStudio.


#### Results

I first calculated the total number of counted cells.
```{r, echo=FALSE}
#Create matrix of initial counts and area
#Also calculate sum of columns
Mus.In.Hypo.Eub.matrix <- matrix(nrow=3, ncol=3)
colnames(Mus.In.Hypo.Eub.matrix) <- c("DAPI","GFP", "Area um2")
Mus.In.Hypo.Eub.matrix[,1] <- c(539, 370, 288)
Mus.In.Hypo.Eub.matrix[,2] <- c(446, 267, 212)
Mus.In.Hypo.Eub.matrix[,3] <- c(4359.63, 2890.21, 1895.51)
MIHE.DAPI <- sum(Mus.In.Hypo.Eub.matrix[,1])
MIHE.GFP <- sum(Mus.In.Hypo.Eub.matrix[,2])
MIHE.Area <- sum(Mus.In.Hypo.Eub.matrix[,3])

Mus.Out.Hypo.Eub.matrix <- matrix(nrow=6, ncol=3)
colnames(Mus.Out.Hypo.Eub.matrix) <- c("DAPI","GFP","Area um2")
Mus.Out.Hypo.Eub.matrix[,1] <- c(107,200,160,182,264,159)
Mus.Out.Hypo.Eub.matrix[,2] <- c(70,123,86,100,169,105)
Mus.Out.Hypo.Eub.matrix[,3] <- c(3183.64,3971.45,3168.87,3980.98,4345.72,3283.65)
MOHE.DAPI <- sum(Mus.Out.Hypo.Eub.matrix[,1])
MOHE.GFP <- sum(Mus.Out.Hypo.Eub.matrix[,2])
MOHE.Area <- sum(Mus.Out.Hypo.Eub.matrix[,3])

Mus.Dep.Hypo.Eub.matrix <- matrix(nrow=4, ncol=3)
colnames(Mus.Dep.Hypo.Eub.matrix) <- c("DAPI","GFP", "Area um2")
Mus.Dep.Hypo.Eub.matrix[,1] <- c(266,316,220,280)
Mus.Dep.Hypo.Eub.matrix[,2] <- c(128,132,137,148)
Mus.Dep.Hypo.Eub.matrix[,3] <- c(3160.2,3466.53,3162.79,3485.15)
MDHE.DAPI <- sum(Mus.Dep.Hypo.Eub.matrix[,1])
MDHE.GFP <- sum(Mus.Dep.Hypo.Eub.matrix[,2])
MDHE.Area <- sum(Mus.Dep.Hypo.Eub.matrix[,3])

Mus.In.Hypo.Ver.matrix <- matrix(nrow=3, ncol=3)
colnames(Mus.In.Hypo.Ver.matrix) <- c("DAPI","GFP", "Area um2")
Mus.In.Hypo.Ver.matrix[,1] <- c(490,396,626)
Mus.In.Hypo.Ver.matrix[,2] <- c(107,80,137)
Mus.In.Hypo.Ver.matrix[,3] <- c(3476.03,3142.46,3939.4)
MIHV.DAPI <- sum(Mus.In.Hypo.Ver.matrix[,1])
MIHV.GFP <- sum(Mus.In.Hypo.Ver.matrix[,2])
MIHV.Area <- sum(Mus.In.Hypo.Ver.matrix[,3])

Mus.Out.Hypo.Ver.matrix <- matrix(nrow=6, ncol=3)
colnames(Mus.Out.Hypo.Ver.matrix) <- c("DAPI","GFP", "Area um2")
Mus.Out.Hypo.Ver.matrix[,1] <- c(145,150,120,226,244,180)
Mus.Out.Hypo.Ver.matrix[,2] <- c(30,21,22,34,51,47)
Mus.Out.Hypo.Ver.matrix[,3] <- c(4328.38,4316.28,4409.73,3971.45,4328.39,4321.46)
MOHV.DAPI <- sum(Mus.Out.Hypo.Ver.matrix[,1])
MOHV.GFP <- sum(Mus.Out.Hypo.Ver.matrix[,2])
MOHV.Area <- sum(Mus.Out.Hypo.Ver.matrix[,3])

Mus.Dep.Hypo.Ver.matrix <- matrix(nrow=5, ncol=3)
colnames(Mus.Dep.Hypo.Ver.matrix) <- c("DAPI","GFP","Area um2")
Mus.Dep.Hypo.Ver.matrix[,1] <- c(264,144,244,304,282)
Mus.Dep.Hypo.Ver.matrix[,2] <- c(29,15,26,41,27)
Mus.Dep.Hypo.Ver.matrix[,3] <- c(2836.98,3787.51,4356.16,4314.53,3611.04)
MDHV.DAPI <- sum(Mus.Dep.Hypo.Ver.matrix[,1])
MDHV.GFP <- sum(Mus.Dep.Hypo.Ver.matrix[,2])
MDHV.Area <- sum(Mus.Dep.Hypo.Ver.matrix[,3])

#Create table to display total counts
TEub.matrix <- matrix(nrow=2,ncol=3)
colnames(TEub.matrix) <- c("Deep","Inlet","Outlet")
rownames(TEub.matrix) <- c("Cells","Bacteria")
TEub.matrix[1,] <- c(MDHE.DAPI, MIHE.DAPI,MOHE.DAPI)
TEub.matrix[2,] <- c(MDHE.GFP, MIHE.GFP,MOHE.GFP)

TVer.matrix <- matrix(nrow=2,ncol=3)
colnames(TVer.matrix) <- c("Deep","Inlet","Outlet")
rownames(TVer.matrix) <- c("Cells","Verrucomicrobia")
TVer.matrix[1,] <- c(MDHV.DAPI, MIHV.DAPI,MOHV.DAPI)
TVer.matrix[2,] <- c(MDHV.GFP, MIHV.GFP,MOHV.GFP)

TEub.matrix

TVer.matrix
```



Using the total counts, the area of the field that I counted, and the volume of water filtered, I determined the number of cells/mL.
```{r, echo=FALSE}
#Convert area from um2 to mm2
MDHE.Area <- MDHE.Area/10^6
MDHV.Area <- MDHV.Area/10^6
MIHE.Area <- MIHE.Area/10^6
MIHV.Area <- MIHV.Area/10^6
MOHE.Area <- MOHE.Area/10^6
MOHV.Area <- MOHV.Area/10^6

#Calculate total filtered area
MDHE.TArea <- 250/MDHE.Area
MDHV.TArea <- 250/MDHV.Area
MIHE.TArea <- 250/MIHE.Area
MIHV.TArea <- 250/MIHV.Area
MOHE.TArea <- 250/MOHE.Area
MOHV.TArea <- 250/MOHV.Area

#Create function to calculate cells/mL
cell.mL <- function(Area,Count){
  cell.filt <- Area * Count
  cells.mL <- cell.filt / 10
  return(cells.mL)
}

#Calculate cells/mL
MDHE.Cell.mL <- cell.mL(MDHE.TArea,MDHE.DAPI)
MDHE.Bac.mL <- cell.mL (MDHE.TArea, MDHE.GFP)
MDHV.Cell.mL <- cell.mL(MDHV.TArea, MDHV.DAPI)
MDHV.Ver.mL <- cell.mL(MDHV.TArea, MDHV.GFP)
MOHE.Cell.mL <- cell.mL (MOHE.TArea, MOHE.DAPI)
MOHE.Bac.mL <- cell.mL(MOHE.TArea, MOHE.GFP)
MOHV.Cell.mL <- cell.mL(MOHV.TArea,MOHV.DAPI)
MOHV.Ver.mL <- cell.mL(MOHV.TArea,MOHV.GFP)
MIHE.Cell.mL <- cell.mL(MIHE.TArea,MIHE.DAPI)
MIHE.Bac.mL <- cell.mL(MIHE.TArea,MIHE.GFP)
MIHV.Cell.mL <- cell.mL(MIHV.TArea,MIHV.DAPI)
MIHV.Ver.mL <- cell.mL(MIHV.TArea,MIHV.GFP)

#Create table to display cells/mL
Eub.matrix <- matrix(nrow=2,ncol=3)
colnames(Eub.matrix) <- c("Deep", "Inlet", "Outlet")
rownames(Eub.matrix) <- c("Cells/mL","Bacteria/mL")
Eub.matrix[1,] <- c(MDHE.Cell.mL,MIHE.Cell.mL,MOHE.Cell.mL)
Eub.matrix[2,] <- c(MDHE.Bac.mL,MIHE.Bac.mL,MOHE.Bac.mL)

Ver.matrix <- matrix(nrow=2,ncol=3)
colnames(Ver.matrix) <- c("Deep", "Inlet", "Outlet")
rownames(Ver.matrix) <- c("Cells/mL","Verrucomicrobia/mL")
Ver.matrix[1,] <- c(MDHV.Cell.mL,MIHV.Cell.mL,MOHV.Cell.mL)
Ver.matrix[2,] <- c(MDHV.Ver.mL,MIHV.Ver.mL,MOHV.Ver.mL)

Eub.matrix

Ver.matrix
```



Next, I calculated the proportion of bacterial and verrucomicrobial cells compared to overall cells.
```{r, echo=FALSE}
#Calculate proporations
DBac.cell <- MDHE.Bac.mL/MDHE.Cell.mL
IBac.cell <- MIHE.Bac.mL/MIHE.Cell.mL
OBac.cell <- MOHE.Bac.mL/MOHE.Cell.mL
DVer.cell <- MDHV.Ver.mL/MDHV.Cell.mL
IVer.cell <- MIHV.Ver.mL/MIHV.Cell.mL
OVer.cell <- MOHV.Ver.mL/MOHV.Cell.mL
DVer.Bac <- MDHV.Ver.mL/MDHE.Bac.mL
IVer.Bac <- MIHV.Ver.mL/MIHE.Bac.mL
OVer.Bac <- MOHV.Ver.mL/MOHE.Bac.mL

#Create proportion table
Prop.matrix <- matrix(nrow=3,ncol=3)
colnames(Prop.matrix)<-c("Deep","Inlet","Outlet")
rownames(Prop.matrix)<-c("Bacteria/Cells","Verruco/Cells","Verruco/Bacteria")
Prop.matrix[1,]<-c(DBac.cell,IBac.cell,OBac.cell)
Prop.matrix[2,]<-c(DVer.cell,IVer.cell,OVer.cell)
Prop.matrix[3,]<-c(DVer.Bac,IVer.Bac,OVer.Bac)

Prop.matrix
```



The proportion of *Verrucomicrobia* compared to Bacteria seemed odd. If Bacteria accounted for 50% of all cells, and *Verrucomicrobia* for 11% of all cells, I would expect the proportion of Verrucomicrobia to Bacteria to be roughly 20%; however, using absolute counts, this is lower at 17%. To better compare the two groups, I calculated the relative proportion of *Verrucomicrobia* with respect to the proportion of Bacteria.
```{r, echo=FALSE}
#Calculate relative proportions
DRVer <- DVer.cell/DBac.cell
IRVer <- IVer.cell/IBac.cell
ORVer <- OVer.cell/OBac.cell

#Create relative proportion table
RelProp.matrix <- matrix(nrow=1,ncol=3)
colnames(RelProp.matrix) <- c("Deep","Inlet","Outlet")
rownames(RelProp.matrix) <- c("Verruco/Bacteria")
RelProp.matrix[1,] <- c(DRVer,IRVer,ORVer)

RelProp.matrix
```



To determine whether the absolute and relative proportions of *Verrucomicrobia* are significantly different, I conducted a test of proportions.
```{r, echo=FALSE}
#Make matrix
Abs.Rel <- matrix(nrow=2,ncol=3)
Abs.Rel[1,] <- Prop.matrix[3,]
Abs.Rel[2,] <- RelProp.matrix

#Transpose matrix; flip dimensions
Abs.Rel <- t(Abs.Rel)

prop.test(Abs.Rel)
```
The p-value is 0.9964, so the relative and absolute proportions of *Verrucomicrobia* are not significantly different. 


For my following calculations, I will use the relative proportion of *Verrucomicrobia*.


To determine whether the *Verrucomicrobia* proportion reported by CARD-FISH differs from the proportion reported by tag sequencing, I conducted another test of proportions.
The sequencing proportions are rounded from preliminary calculations performed by the Denef Lab.

```{r, echo=FALSE}
#Make sequencing proportion table
Tag.matrix <- matrix(nrow=1, ncol=3)
colnames(Tag.matrix) <- c("Deep", "Inlet", "Outlet")
Tag.matrix[1,] <- c(0.31256, 0.09375,0.21875)

#Create table for comparison of seq and rel CARD-FISH
Tag.compare <- matrix(nrow=3, ncol=2)
Tag.compare[,1] <- Abs.Rel[,2]
Tag.compare[,2] <- c(0.31256, 0.09375,0.21875)

prop.test(Tag.compare)
```


To help visualize the data, I plot the CARD-FISH and tag sequencing proportions with error bars.
```{r, echo=FALSE}
#Function to calculate standard deviation
sd <- function(p,n){
  a <- 1-p
  b <- p*a
  c <- b/n
  d <- sqrt(c)
  return(d)
}

#Calculate sd
FISH.Dep.sd <- sd(Abs.Rel[1,2], 1000)
FISH.In.sd <- sd(Abs.Rel[2,2], 1000)
FISH.Out.sd <- sd(Abs.Rel[3,2], 1000)
Tag.Dep.sd <- sd(Tag.compare[1,1], 1000)
Tag.In.sd <- sd(Tag.compare[2,1], 1000)
Tag.Out.sd <- sd(Tag.compare[3,1],1000)

#Prep matrix for bar plot
Tag.compare <- t(Tag.compare)

#Create bar plot
barplot(Tag.compare, width=1, names.arg=c("Deep", "Inlet", "Outlet"),beside=TRUE, horiz=FALSE, col=c("Yellow","Blue"),main="Verrucomicrobia Proportion", xlab="Sample", ylab="Relative Proportion", xlim=c(0,17),ylim=c(0,0.35), axisnames=TRUE)

#Add legend
legend(x=c(10,17), y=c(0,0.1), legend=c("CARD-FISH","Sequencing"), col=c("yellow","blue"), pch=15)

#Add std error bars
#Vertical
segments(1.5,Tag.compare[1,1]-FISH.Dep.sd,1.5,Tag.compare[1,1]+FISH.Dep.sd,lwd=2)
#Bottom horizontal
segments(1.25,Tag.compare[1,1]-FISH.Dep.sd, 1.75, Tag.compare[1,1]-FISH.Dep.sd,lwd=2)
#Top horizontal
segments(1.25,Tag.compare[1,1]+FISH.Dep.sd, 1.75, Tag.compare[1,1]+FISH.Dep.sd,lwd=2)

segments(2.5,Tag.compare[2,1]-Tag.Dep.sd,2.5,Tag.compare[2,1]+Tag.Dep.sd,lwd=2)
segments(2.25,Tag.compare[2,1]-Tag.Dep.sd, 2.75, Tag.compare[2,1]-Tag.Dep.sd,lwd=2)
segments(2.25,Tag.compare[2,1]+Tag.Dep.sd, 2.75, Tag.compare[2,1]+Tag.Dep.sd,lwd=2)

segments(4.5,Tag.compare[1,2]-FISH.In.sd,4.5,Tag.compare[1,2]+FISH.In.sd,lwd=2)
segments(4.25,Tag.compare[1,2]-FISH.In.sd,4.75,Tag.compare[1,2]-FISH.In.sd,lwd=2)
segments(4.25,Tag.compare[1,2]+FISH.In.sd,4.75,Tag.compare[1,2]+FISH.In.sd,lwd=2)

segments(5.5,Tag.compare[2,2]-Tag.In.sd,5.5,Tag.compare[2,2]+Tag.In.sd,lwd=2)
segments(5.25,Tag.compare[2,2]-Tag.In.sd,5.75,Tag.compare[2,2]-Tag.In.sd,lwd=2)
segments(5.25,Tag.compare[2,2]+Tag.In.sd,5.75,Tag.compare[2,2]+Tag.In.sd,lwd=2)

segments(7.5,Tag.compare[1,3]-FISH.Out.sd,7.5,Tag.compare[1,3]+FISH.Out.sd,lwd=2)
segments(7.25,Tag.compare[1,3]-FISH.Out.sd,7.75,Tag.compare[1,3]-FISH.Out.sd,lwd=2)
segments(7.25,Tag.compare[1,3]+FISH.Out.sd,7.75,Tag.compare[1,3]+FISH.Out.sd,lwd=2)

segments(8.5,Tag.compare[2,3]-Tag.Out.sd,8.5,Tag.compare[2,3]+Tag.Out.sd,lwd=2)
segments(8.25,Tag.compare[2,3]-Tag.Out.sd,8.75,Tag.compare[2,3]-Tag.Out.sd,lwd=2)
segments(8.25,Tag.compare[2,3]+Tag.Out.sd,8.75,Tag.compare[2,3]+Tag.Out.sd,lwd=2)
```
The p-value is 0.9508, so the proportions of *Verrucomicroba* reported by CARD-FISH and tag sequencing are not significantly different.


#### Discussion
It would be interesting to see if probes targeted towards gram-positive bacteria performed similarly despite the added challenge of permeating the thick peptidoglycan layer. I would also be interested in seeing if more specific probes--such as those targeting a bacterial class--perform similarly.

For future CARD-FISH runs, I would like to examine specific classes of *Verrucomicrobia*.

Most importantly, I need to finish learning how to analyze the OTU data exported from Mothur. This will be the basis of my senior thesis and will depend upon statistical analysis with R. In other words, I need to step up my game.




